Assessing the Accessibility of Protein Tunnels using Motion Planning

### 2017-12-09 paper.tex
table 2 regeneratd, new type of graph in Fig, 2

The aim of this experiment was to verify the procedure to select feasible ligand conformations for the motion planning task.
The experiment was performed with the ligand DCP (Dichloropropanol), for which $1300$ conformations were available, and for the ligand
m040 (1,5-Dichloropentane) with $7560$ available conformations.
Two strategies were tested: the strategy described in Sec.~\ref{sec::strat} (denoted `S') and the strategy that prefers smaller conformations (denoted `P').
The `P' method selects such conformations where the deviations of the atomic distance from the geometric center are the lowest, which leads 
to the selection of packed conformations (examples are shown in Tab.~\ref{fig::m040c}).

The traversability rates are shown in Tab.~\ref{tab::selection}, where the subscript of each ligand's 
name shows the number of the selected conformations (5 or 50) and the used method (S or P).
The number after the slash in the ligand's name denotes the number of atoms of the ligand.
In the case of both tested ligands and 50 selected conformations, the `S' strategy leads to a higher traversability rate than the `P' strategy.
For the DCP ligand, a higher traversability rate was achieved when 50 conformations were selected.
This experiment shows that selecting 50 conformations using the `S' strategy (Sec.~\ref{sec::strat}) leads to better results
than using a less number of conformations or when preferring the packed conformations.

\begin{table}[ht]
\centering
\caption{\label{tab::selection}
    \small
    The influence of the conformation selection method on the traversability rates in the first (main) tunnel.
    The number after~`$/$' denotes the number of atoms.
    The protein has 4650 atoms.
}
\small
\renewcommand{\tabcolsep}{4.3pt}
{\scriptsize
\renewcommand{\arraystretch}{0.7}
\input graphSel.tex
}
\end{table}

@@

Based on the results of the previous experiment, 50 conformations using the `S' strategy were selected for ligands m037t (1,2-Dichloroethane), m038t (1,3-Dichloropropane), m056 (2-Chlorobutane), and m080 (Trichloropropane).
Models of the ligands in Mol2 formats were converted to param files~\cite{meiler2006rosettaligand} containing the topology, rotatable bonds, atom types and partial charges. 
Rotamer library was created by rotation of all rotatable bonds with 30$^\circ$ step using Rosetta3.
The energy cut-off for individual conformations was set to 0.075. %(-rot_ensemble_ecutoff 0.075) 
%and Rosetta database 3.2.1 was used during the conformers generation.

%\cite{meiler2006rosettaligand}
%All substrates were prepared by XXXXXX program and saved in mol2 file. Mol2 formats were converted to param files [1] containing the topology, rotatable bonds, atom types and partial charges by molfile_to_params.py script supplemented in Rosetta Scripts [2].

%Rotamer library was created by rotation of all rotatable bonds with 30$^{\circ}$ step. 
%The energy cut-off for individual conformations was set to 0.075. % (-rot_ensemble_ecutoff 0.075) 
%and Rosetta database 3.2.1 was used during the conformers generation.
%Conformations passing the energy criteria were output as multi-pdb file.

@@

The traversability rate is shown in Tab.~\ref{tab::main}. 
The table shows the traversability rate only for the main tunnel, as no trajectory was found in the side tunnel for most of the tested ligands.
The only exception is the ligand m037t, which was the smallest tested one (with 8 atoms), and therefore it could fit even inside the side tunnel.
The traversability rate is higher for the most scaled-down ligands ($\smin=0.5$) and it decreases with the increasing scale.
For the ligand DCP, trajectories were detected in 10~\% of frames for scale $\smin=0.8$ and the traversability of this ligand
to the active site was approved by a separate MD simulation~\cite{marques2017catalytic}.
%For the scale $\smin=0.8$, the most traversable frames were detected for ligands DCP (18~\%), m037t (45~\%) and m038t (30~\%).
%The traversability of the DCP ligand to active site was approved by a separate MD simulation~\cite{marques2017catalytic}.
Based on the number of success rate, also other ligands (e.g., m037t with traversability rate 45~\% at $\smin$=0.8) seem to be promising.
However, their binding was not approved yet by MD dedicated simulations.

Beside scaling-down the ligand, also the atoms of the protein can be scaled-down, which is used e.g. in~\cite{cortes2005path} which
scales  atoms of both the ligand and the protein to 80~\% of their size.
We tested also this case (column \RD\ in Tab.~\ref{tab::main}).
The results are same as for ligand scale $\smin=0.6$ (except DCP and m040 which can be result of the randomization).

@@

%originally last two paragraphs in the experiment section, for resubmition, they are moved here to discussion
The proposed \RA\ planner provides higher traversability rate than RRT-Retraction (\RB).
The distances of the configuration trees (measured using the 3D Euclidean metric) to the active site and the runtimes in each frame
are shown in Fig.~\ref{fig::comparison}.
The runtimes of both algorithms increase in those frames, where the tunnel is not traversable, i.e., where the distance
to the active site is more than 2~\AA\ (for example, in the frame no. 60).
The comparison between \RA\ and \RC\ shows that utilization of the  atom-based metric (used in \RA) leads to higher traversability rate
in the case of large scales and larger ligands (e.g., m056 with $\smin=0.8$: 15~\% with the metric, but to only 6~\% without it).
Both \RA\ and \RC\ provided trajectories with the same traversability rate in the case of lower scales ($\smin<0.6$).

>>>>

The aim of this experiment was to verify the procedure to select feasible ligand conformations from an available pool of conformations.
Models  of the ligands in Mol2 formats were converted to param files~\cite{meiler2006rosettaligand} containing the topology, rotatable bonds, atom types and partial charges. 
The pool of conformations $\L$ was created by rotation of all rotatable bonds with 30$^\circ$ step using Rosetta3.
The energy cut-off for individual conformations was set to 0.075. %(-rot_ensemble_ecutoff 0.075) 
Two strategies to select a subset from the $\L$ conformations were tested: 
the strategy described in Sec.~\ref{sec::strat} (denoted `S') and the strategy that prefers smaller conformations (denoted `P').
The `P' method selects such conformations where the deviations of the atomic distance from the geometric center are the lowest, which leads 
to the selection of packed conformations.
The strategies were tested using the ligand DCP (Dichloropropanol), for which $1300$ conformations were created, and for the ligand
m040 (1,5-Dichloropentane) with $7560$ conformations.

The traversability rates are shown in Tab.~\ref{tab::selection}, where the subscript of each ligand's 
name shows the number of the selected conformations (5 or 50) and the used method (S or P).
The number after the slash in the ligand's name denotes the number of atoms of the ligand.
In the case of both tested ligands and 50 selected conformations, the `S' strategy leads to a higher traversability rate than the `P' strategy.
For the DCP ligand, higher traversability rate was achieved when 50 conformations were selected.
This experiment shows that selecting 50 conformations using the `S' strategy (Sec.~\ref{sec::strat}) leads to better results
than using a less number of conformations or when preferring the packed conformations.

\begin{table}[ht]
\centering
\caption{\label{tab::selection}
    \small
    The influence of the conformation selection method on the traversability rates in the first (main) tunnel.
    The number after~`$/$' denotes the number of atoms.
    The protein has 4650 atoms.
}
\small
\renewcommand{\tabcolsep}{4.3pt}
{\scriptsize
\renewcommand{\arraystretch}{0.7}
\input graphSelNew.tex
}
\end{table}

@@

Based on the results of the previous experiment, 50 conformations using the `S' strategy were selected for ligands 
m037t (1,2-Dichloroethane), m038t (1,3-Dichloropropane), m056 (2-Chlorobutane), and m080 (Trichloropropane).

    %and Rosetta database 3.2.1 was used during the conformers generation. 
%\cite{meiler2006rosettaligand}
%All substrates were prepared by XXXXXX program and saved in mol2 file. Mol2 formats were converted to param files [1] containing the topology, rotatable bonds, atom types and partial charges by molfile_to_params.py script supplemented in Rosetta Scripts [2]. 
%Rotamer library was created by rotation of all rotatable bonds with 30$^{\circ}$ step. 
%The energy cut-off for individual conformations was set to 0.075. % (-rot_ensemble_ecutoff 0.075) 
%and Rosetta database 3.2.1 was used during the conformers generation.
%Conformations passing the energy criteria were output as multi-pdb file.

@@

The traversability rate is shown in Tab.~\ref{tab::main}. 
The table shows the traversability rate only for the main tunnel, as no trajectory was found in the side tunnel for most of the tested ligands.
The only exception is the ligand m037t, which was the smallest tested one (with 8 atoms), and therefore it could fit even inside the side tunnel.
The traversability rate is higher for the most scaled-down ligands ($\smin=0.5$) and it decreases with the increasing scale.
For the ligand DCP, trajectories were detected in 18~\% of frames for scale $\smin=0.8$ and the traversability of this ligand
to the active site was approved by a separate MD simulation~\cite{marques2017catalytic}.
%For the scale $\smin=0.8$, the most traversable frames were detected for ligands DCP (18~\%), m037t (45~\%) and m038t (30~\%).
%The traversability of the DCP ligand to active site was approved by a separate MD simulation~\cite{marques2017catalytic}.
Based on the number of success rate, also other ligands (e.g., m037t with traversability rate 45~\% at $\smin$=0.8) seem to be promising.
However, their binding was not approved yet by MD dedicated simulations.

Beside scaling-down the ligand, also the atoms of the protein can be scaled, which is used e.g. in~\cite{cortes2005path}.
The tool~\cite{cortes2005path} uses the scale 0.8 for both the ligand and the protein.
We tested also this case (the column \RD\ in Tab.~\ref{tab::main}).
The results are same as for ligand scale $\smin=0.6$ (except DCP and m040 which can be a result of the randomization).

@@

%originally last two paragraphs in the experiment section, for resubmition, they are moved here to discussion
The proposed \RA\ planner provides higher traversability rate than RRT-Retraction (\RB).
The distances of the configuration trees (measured using the 3D Euclidean metric) to the active site and the runtimes in each frame
are shown in Fig.~\ref{fig::comparison}.
The runtimes of \RB\ are higher than runtimes of \RA and the std. deviations as well.
The variant \RC\ is faster than \RA\ due to simple evaluation of the metric in the expansion step.
%The runtimes of both algorithms increase in those frames, where the tunnel is not traversable, i.e., where the distance
%to the active site is more than 2~\AA\ (for example, in the frame no. 60).
The utilization of the  atom-based metric (used in \RA) leads to higher traversability rate
in the case of large scales and larger ligands (e.g., m056 with $\smin=0.8$: 15~\% with the metric, but to only 6~\% without it).
Both \RA\ and \RC\ provided trajectories with the same traversability rate in the case of lower scales ($\smin<0.6$).
